Brief Article

### 2012-02-19 power.tex
fixed the 1- bug
two methods of computing pB

Let us think of a lineup as a head-to-head comparison of the data plot and $m-1$ null plots. Let the data plot have a $p$ value of $p_B$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
We know that for each comparison the probability that the data plot has the smaller $p$ value is 
\begin{eqnarray*}
P(p_B < p_{0,i}) &=& 1 - P(p_{i,0} \le p_B) = 1- \int_0^1  P(p_{i,0} \le p_B \mid p_B=t) f_{p_B}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_B}(t) dt = 1 - \int_0^1 t f_{p_B}(t) dt = 1 - E[p_B],
\end{eqnarray*}
which, in particular, is independent of $p_{0,i}$ for all $i$.

Let us now define $X$ as the number of null plots in a lineup, for which the $p$ value $p_{0,i}$ is smaller than $p_B$ (i.e. under the assumption that the onlooker is able to identify the chart with the smallest $p$ value, they would pick any of the $X$  null plots over the data plot). 
Then $X \sim B_{m-1, E[p_B]}$, and the probability that an onlooker will pick the data plot in a given lineup is 
\[
P(X=0) = \left(1 - E[p_B] \right)^{m-1} = P(p_B \le p_0),
\]
where $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$

\section{Expected number of choices}
Since $X$ has a binomial distribution, we can have a look at the number of plots among the null plots that we would expect to be picked over the data plot:
\[
E[X] = (m-1)(1-E[p_B]).
\]
With an increase of the lineup size $m$ the number of null plots with a potentially stronger signal than the data plot grows linearly.
This should allow us to infer some of the signal strength $p_B$ as
\[
E[p_B] = 1 - E[X]/(m-1),
\]
i.e. by averaging over the number of plots with  a stronger signal than the data plot we can evaluate signal strength of the plot even in the case where the data plot is not being picked by the observer.

Now assume that the same lineup plot is evaluated by $K$ independent observers. Wlog let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of people who picked plot $i$ as the plot with the strongest expression of non-random structure, then $\sum_{k=1}^{m} n_k = K$, and $(n_1, n_2, ..., n_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{images/powerplot.pdf} 
   \caption{Comparison of probabilities to pick the data plot for different size lineups $m$ and different values for data plot strength $p_B$. The grey line shows probability of picking the data plot under $H_0$. We see that for a $p$ value $p_B$ of about 0.15 for the dataplot the signal in the plot is so weak, that we cannot distinguish the dataplot from nullplots in a lineup of size $m=20$. (Mahbub, is that what we see in practice?)}
   \label{power}
\end{figure}
Based on $K$ observers, we can assume that $n_m$ votes for the dataplot represent support for the data over any of the nullplots, i.e. $\pi_m = P(X=0)$.

Under $H_1$ we therefore get an estimate for the strength of the dataplot signal $p_B$ as
\[
\widehat{E[p_B]} = 1 - \left( \frac{n_m}{K} \right)^{1/(m-1)}.
\]
%Similarly, we get an estimate for $E[X]$, the average number of null plots with a stronger signal strength than the data plot as the number of nullplots in the lineup with more than $n_m$ picks in their favor, which leads us to another estimate of $E[p_B]$ as 
%\[
%\widehat{E[p_B]}  = 1 - 1/(m-1) \cdot \sum_i I (n_i > n_m),
%\]
%where $I(x) = 1$ for $x$ is true and $I(x) = 0$ if $x$ is false.
\end{document}

>>>>

Let us think of a lineup as a head-to-head comparison of the data plot and $m-1$ null plots. Let the data plot have a $p$-value of $p_B$, and the null plots $p$-values of $p_{0, i}$ with $1 \le i \le m-1$.
We know that for each comparison the probability that the data plot has the smaller $p$-value is 
\begin{eqnarray*}
P(p_B < p_{0,i}) &=& 1 - P(p_{i,0} \le p_B) = 1- \int_0^1  P(p_{i,0} \le p_B \mid p_B=t) f_{p_B}(t) dt =  \\
&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_B}(t) dt = 1 - \int_0^1 t f_{p_B}(t) dt = 1 - E[p_B],
\end{eqnarray*}
which, in particular, is independent of $p_{0,i}$ for all $i$.

Let us now define $X$ as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_B$ (i.e. under the assumption that the onlooker is able to identify the chart with the smallest $p$-value, they would pick any of the $X$  null plots over the data plot). 
Then $X \sim B_{m-1, E[p_B]}$, and the probability that an onlooker will pick the data plot in a given lineup is 
\begin{equation}\label{p0}
P(X=0) = \left(1 - E[p_B] \right)^{m-1} = P(p_B \le p_0),
\end{equation}
where $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$

\section{Expected number of choices}
Since $X$ has a binomial distribution, we can have a look at the number of plots among the null plots that we would expect to be picked over the data plot:
\[
E[X] = (m-1)E[p_B].
\]
With an increase of the lineup size $m$ the number of null plots with a potentially stronger signal than the data plot grows linearly.
This allows us to infer some of the signal strength $p_B$ as
\[
E[p_B] = E[X]/(m-1),
\]
i.e. by averaging over the number of plots with  a stronger signal than the data plot we can evaluate signal strength of the plot even in the case where the data plot is not being picked by the observer:

assume that the same lineup plot is evaluated by $K$ independent observers. Wlog let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of people who picked plot $i$ as the plot with the strongest expression of non-random structure, then $\sum_{k=1}^{m} n_k = K$, and $(n_1, n_2, ..., n_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

We can then find an estimate for $X$, the number of null plots with a stronger signal than the data plot, as the number of null plots with more votes $n_i$ than the dataplot $n_m$. This gives an estimate for $X$ as
\[
\hat{X} = \sum_{i=0}^{m-1} I(n_i \ge n_m),
\]
where $I(x) = 1$ if $x$ is true and 0 otherwise.

However, this is not a very precise estimator for $E[p_B]$, since the only possible values for the  estimates are multiples of $1/(m-1)$.

As an alternative, we will make use of the fact that $\pi_m = P(X=0)$, since a vote for the dataplot implies that none of the other plots have a stronger signal.

With $\widehat{\pi_m} = n_m/K$ as an estimate for $\pi_m$, we get an plugin estimate for $E[p_B]$ from eq (\ref{p0}) as
\[
\widehat{E[p_B]} = 1 - \left( \frac{n_m}{K} \right)^{1/(m-1)}.
\]

Since $n_m$ has a binomial distribution $n_m \sim B_{m-1, \pi_m}$, we can also derive a variance of the above estimate as
\[
Var \ \widehat{p_B} = K^{-2/(m-1)} Var \left[n_m^{1/(m-1)}\right]
\]

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{images/powerplot.pdf} 
   \caption{Comparison of probabilities to pick the data plot for different size lineups $m$ and different values for data plot strength $p_B$. The grey line shows probability of picking the data plot under $H_0$. We see that for a $p$-value $p_B$ of about 0.15 for the dataplot the signal in the plot is so weak, that we cannot distinguish the dataplot from nullplots in a lineup of size $m=20$. }
   \label{power}
\end{figure}
%Based on $K$ observers, we can assume that $n_m$ votes for the dataplot represent support for the data over any of the nullplots, i.e. $\pi_m = P(X=0)$.
%
%Under $H_1$ we therefore get an estimate for the strength of the dataplot signal $p_B$ as
%\[
%\widehat{E[p_B]} = 1 - \left( \frac{n_m}{K} \right)^{1/(m-1)}.
%\]
%Similarly, we get an estimate for $E[X]$, the average number of null plots with a stronger signal strength than the data plot as the number of nullplots in the lineup with more than $n_m$ picks in their favor, which leads us to another estimate of $E[p_B]$ as 
%\[
%\widehat{E[p_B]}  = 1 - 1/(m-1) \cdot \sum_i I (n_i > n_m),
%\]
%where $I(x) = 1$ for $x$ is true and $I(x) = 0$ if $x$ is false.
\end{document}
